---
title: "Most Important Fisheries - Brazil"
author: "Daniel Viana"
date: "October 4, 2018"
output:   
  html_document:
    toc: yes
    toc_float: yes
    pdf_document: default
  word_document:
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(tidyr)
library(ggplot2)
library(kableExtra)
library(data.table)
library(knitr)

ggtheme_plot <- function() {
  theme(
    plot.title = element_text(size = rel(1.25), hjust = 0, face = "bold"),
    panel.background = element_blank(),
    strip.background = element_blank(),
    # strip.text       = element_text(size = base_size, face = "italic"),
    panel.border     = element_blank(),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_blank(),
    axis.ticks = element_blank(),
    axis.text.x = element_text(size = 16),
    axis.text.y = element_text(size = 16),
    axis.title = element_text(size = 16),
    legend.key = element_rect(colour = NA, fill = NA),
    legend.position  = "top",
    legend.title = element_text(size = 18),
    legend.text = element_text(size = 16),
    strip.text.x = element_text(size = 20, colour = "darkgrey")
  )
}


```

#1. Introduction

Brazil has over 7 thousand kilometers of coastline and one of the largest river networks in the world, with only the Amazon River covering over 7 million kilometers square of water. Fisheries resources in Brazil are extremely important for income and food security of millions of people throughout the country. Despite enormous social and economic importance, many fish stock are over exploited and in risk of collapse. According to a study conducted by the Ministry of Environment, 80% of fish stocks are over exploited or fully exploited. This signals the urgency for sustainable management of fisheries in order to guarantee the livelihoods of fishers, improve food security in the country, and conserve marine biodiversity. 

There is no basic fisheries statistics in the country since 2008. The lack of national fisheries statistics posed a great challenge to this study. Several researchers and fisheries managers consulted during the study pointed out the lack of data to provide basic information on Brazilian fisheries. Information such as how much the country produces, how much it consumes or even the number of fishers in the country is very sparse. In an interview with the most renowned fisheries scientist in the country, Dr. Fabio Hazin, he pointed out that there are between 400 thousand and one million fishers in the country. Additionally, he pointed out that data from FAO on Brazilian fisheries is mostly a copy of the production from the previous years. And this has been going on since 2008, when the last year of national fisheries statistics was recorded, and even then it wasn't very reliable. Therefore, we had to be creative to conduct this study. We contacted people from all fisheries sectors to find out what was available. After an extensive search, we gathered six different databases representing the different fisheries sectors and conducted an extensive literature research. We also identified other potential sources of information that we were not able to obtain in time for this study. 

The databases used were:

1. SIGSIF - Data obtained from the sanitary inspection services of the Brazilian Ministry of Agriculture. This database contains all seafood processed by fishing industries that caries a federal inspection certificate. This certificate is required for industries to commercialize their product to different states of Brazil and to export and import seafood products. It is a rigorous inspection program, where one of the requirements is to report all seafood that enters each company. Thus, this database includes all seafood products by species imported or from domestic production that enters all seafood processing companies with the federal inspection certificate (total of 472 companies). It includes both aquaculture and wild caught products from both freshwater and the oceanic species. 2016-2017. 

2. CEAGESP - CEAGESP is the largest wholesale open seafood market in Latin America, located in the city of São Paulo. It is run by a state-owned company where sellers and buyers from all over the country gather to commercialize seafood and seafood products. We obtained information on all the seafood products commercialized in the market by species and price of commercialization. 2016-2017.

3. MDIC - Import and Export data from the Ministry of Development, Industry and Foreign Trade. This database contains all seafood imported and exported in Brazil. 2014-2017.

4. State fisheries landings - Fisheries landings data from three states of Brazil were there is continuous data collection and data are publicly available. These three states (São Paulo, Parana and Santa Catarina) are located in Southern Brazil and have a continuous data collection program sponsored by Petrobras in partnership with Universities. Petrobras is required to collect landings information for insurance purposes and has made some of its data publicly available for research projects. 2016-2017.

5. IBGE - Brazilian Institute of Geography and Statistics. This federal institute is responsible for surveying the country to provide basic information on all social aspects of Brazil. We gathered two pieces of information to inform our analysis: household food consumption and income by region of Brazil (last survey dates from 2009 and 2010). 

6. Fishing Permits - Number of registered fishers in the country from the extinct Ministry of Fisheries and Aquaculture (MPA). 2003 - 2010.


#2. Fisheries Context

FAO and SAU data

#3. Fisheries Employment and Income 

## Top Fisheries by Employment

As pointed out in the introduction, Brazil has no reliable data on the number of fishers in the country. To respond this question we put together several pieces of information from different sources to have an estimate of the importance of fisheries for employment in Brazil. 

Data from IBGE (2013) estimates that Brazil had about 500 thousand fishers throughout the country, with 90% from the small-scale sector and 10% from the industrial sector (Campos and Chaves 2016). Data from the extinct Ministry of Fisheries and Aquaculture (MPA) suggests there are many more fishers in Brazil (Figure 1). According to these data, there were about 900 thousand registered fishers in 2010 throughout the country, with 95% of the registered fishers from the small-scale sector (MPA, 2013). There is a large discrepancy between these two sources, which reiterates the uncertainty around estimates of the number of fishers in the country. Nonetheless, these numbers highlight the huge importance of the fisheries sector to employment in Brazil. 

```{r}
###Number of registered fishers
rgp.data = read.csv("rgp_MPA.csv")

ggplot(rgp.data, aes(x = year, y = fishers)) + 
  geom_bar(stat = "identity") + 
  labs(x="Year", y="Number of Registered Fishers", caption = "Figure 1 - Number of registered fishers in Brazil from 2003 to 2010 (Source: Minitry of Fisheries and Aquaculture)") + 
  ggtheme_plot()

```

Employment along the supply chain varies according to the region, fisheries and species commercialized. In the artisanal sector, most of the catch is consumed regionally, or in nearby capitals. However, some valuable fish and seafood are also commercialized in the main capitals - São Paulo and Rio de Janeiro. Typically, there are three main supply chain configurations in artisanal fisheries. First, there are estuarine/mangrove species (mostly crabs, shrimp and bivalves) that are commercialized whole in local markets or processed to achieve larger markets. For example, processing of crab meat is an important activity for coastal communities, especially for fisher woman. To obtain one kilogram of crab meat, a full day's work is required, which involves extraction of the crab, cooking and manually removing the meat. This meat is then distributed through middleman to nearby cities and capitals. Second, most of the fish captured by the artisanal sector is consumed locally or regionally. In most cases, there is no record of the catch or commercialization of these fish, being sold in local markets and restaurants. Third, more valuable fish such as snappers, groupers, snook, weakfish and large shrimp are sold in capitals throughout the country. These fish are sold through middleman and larger seafood industries to large capitals, thus involving a larger number of people in the supply chain. Again, there is no reliable estimate on the total number of people employed in the supply chain of artisanal fisheries. The MPA estimates that there are about two million people involved in these fisheries (MPA, 2013). 

Assuming that the number of fishers are between 500 thousand and 1 million, and that about 90% of them are from the small-scale sector, the most important fisheries for employment in the country are those from the artisanal sector. The artisanal sector is where we have the least amount of data on the production and species harvested. There is an estimate from the MPA that 50-70% of the catch from the entire country comes from the artisanal sector (MPA, 2013). Small-scale fisheries in Brazil target hundreds of species and there is no reliable information on the most important species for this sector. 

Most of the registered fishers in Brazil are from the North and Northeast regions (MPA report, 2013). In these regions, freshwater resources from the Amazon and estuarine fisheries are very important. From the Amazonas state, the top commercialized fish are: curimatã (Prochilodus nigricans), jaraqui (Semaprochilodus spp.), matrinchã (Brycon amazonicus), pacu (Mylossoma duriventre), tambaqui (Colossoma macropomum) and tucunaré (Cichla monoculus) (data from the secretary of fisheries in the Amazon state - Sepror, 2018). In estuaries, there are many important resources such as the mangrove crabs, shrimps, bivalves, weakfish and snook (information gathered though literature review and interviews). Besides these fisheries, the small-scale fishers target all ecosystems and resources, such as pelagic, reefs, demersal and coastal. In the Northeast region for example, reef fisheries are very important, capturing some highly valued species such as snappers, groupers and lobsters. In the south, the mullet fishery is very important, with the first quota system for Brazil implemented in 2018 to manage this fishery. Therefore, the artisanal sector is a multi-species fishery, with the main species depending on the region and ecosystem. 

Industrial fisheries in Brazil employ only about 10% of the fishers and is responsible for about 30-50% of the fisheries production. The industrial sector in Brazil is mainly concentrated in the south of the country, especially in Santa Catarina state where the main industrial fisheries port is located (Itajai port). Landings data from Santa Catarina indicate that sardines are the most important species, followed by corvina and skipjack tuna (Table 1). Another important industrial fisheries fleet is located in the mouth of the Amazon river, in the north of Brazil. The main resource captured in the region is the catfish piramutaba (Brachyplatystoma vaillantii), although there is no data available from this fleet. Therefore, the top fisheries by employment in the industrial sector are represented in the landings data from the Santa Catarina state and the piramutaba fishery in the north.  

```{r}
#State of Santa Catarina
sc.data = read.csv("Santa Catarina.csv")

###Graph main species - production
sc.prod = sc.data %>% 
  filter(year == "2017") %>%
  group_by(species_english) %>%
  summarize(sum(production_kg)) %>%
  rename(Species = species_english, Production = "sum(production_kg)") %>%
  arrange(desc(Production))%>%
  transform(Percent = Production/sum(Production)*100)

  kable(sc.prod[1:10,], caption = "Table 1 - Main species caught in Santa Catarina state, Brazil in 2017 (Source: Univali)", booktabs = T) %>%
  kable_styling(latex_options = c("striped", "hold_position"))

```

In the industrial sector, the majority of the catch is processed by large companies to be distributed in supermarkets and restaurants all over the country. According to data published by the Seafood Industry Association (ABIPESCA), the seafood industry employs about 8 thousand direct employments and 30 thousand indirect employments. Data from the seafood companies (SIGSIF) indicates that the most important industrial fisheries are sardines, croacker, tuna and piramutaba catfish (Table 2). Thus, the industrial sector employs relatively less people per ton of captured fish compared to the artisanal sector. 


```{r}
data.sigsif = read.csv("SIGSIF.csv")


###Rank main species comercialized 
domestic.fisheries = data.sigsif %>% 
  filter(origin == "PRODUTOR" & form == "EXTRATIVO" & year == "2017") %>%
  group_by(species_english) %>%
  summarize(sum(quantity)) %>%
  rename(Species = species_english, Quantity = "sum(quantity)") %>%
  arrange(desc(Quantity)) %>%
  transform(Percent = Quantity/sum(Quantity)*100)  


  kable(domestic.fisheries[1:10,], caption = "Table 2 - Main species comercialized by seafood industries with SIF certificate in Brazil (Source: Ministry of Agriculture)", booktabs = T) %>%
  kable_styling(latex_options = c("striped", "hold_position"))


```

## Top Fisheries by Income

Because there is no reliable national-level landings data in Brazil since 2008, it is challenging to determine the top fisheries by income generated in the country. Data available are mainly from the industrial sector, where sardines, croaker and tuna are the main species caught. However, information from the SIGSIF (Table 2) and Santa Catarina state landings does not have any information on price. 

A report by Nielsen analytic (Nielsen, 2018) shows that the canned sardine market in Brazil is valued at about 1.4 billion Brazilian Reais (approx. 380 million Dollars) and the canned tuna market is worth about 145 million Brazilian Reais (approx. 39 million Dollars). Additionally, according to the Brazilian seafood industry association, seafood industries have an aggregate revenue of 8 billion Brazilian Reais (approx. 2.2 billion Dollars) per year. These numbers demonstrate the strength of the seafood market (especially canned market) in Brazil and the importance of sardines for income and food security in the country.

Data from SEAGESP market have important information on the value of seafood in the country. This database contains quantity and price information for all seafood products. Products sold in the market are both from industrial and artisanal sectors. Table 3 shows the main species ranked by value sold in the CEAGESP market in 2017. The most valuable species in this year were sardines followed by sharks, croaker, tuna and acoupa weakfish (Table 3). Sharks are not distinguished by species and are captured by both artisanal and industrial fleets. Sardines, croakers and tunas are caught mainly by the industrial fleet, while the acoupa weakfish is mainly caught by artisanal fisheries from the north and northeast regions of Brazil. Snook is also one of the top 10 species sold in the market and mainly captured by the artisanal sector. It is one of the species with greater price in the market.    

```{r}
ceagesp.data = read.csv("producao_seagesp.csv")

###Graph main species - production
ceagesp.value = ceagesp.data %>% 
  filter(year == "2017" & origin == "extraction") %>%
  group_by(species_english) %>%
  summarize(sum(value_rs)) %>%
  rename(Species = species_english, Value_RS = "sum(value_rs)") %>%
  arrange(desc(Value_RS)) %>%
  transform(Percent = Value_RS/sum(Value_RS)*100) 
  
  kable(ceagesp.value[1:10,], caption = "Table 3 - Main species comercialized in 2017 at CEAGESP fishmarket, Sao Paulo, Brazil (Source: CEAGESP)", booktabs = T) %>%
  kable_styling(latex_options = c("striped", "hold_position"))

```

Many important small-scale resources are not sold in the SEAGESP market but are important sources of income through sale in local markets. Resources such as crabs, mullets, shrimp, snappers and groupers are very important sources of income for small-scale fishers. Only from our databases there are over 500 different species reported. For example, in Paraná state most of the production is from small-scale fisheries sector. In this state, shrimps, weakfishes and crab are the most important species in terms of value (Table 4). 

```{r}
#State of Parana
pa.data = read.csv("parana.csv")

###Rank main species - value
pa.value = pa.data %>% 
  filter(year == "2016") %>%
  group_by(species_english) %>%
  summarize(sum(value_rs)) %>%
  rename(Species = species_english, Value_RS = "sum(value_rs)") %>%
  arrange(desc(Value_RS)) %>%
  transform(Percent = Value_RS/sum(Value_RS)*100) 

  kable(pa.value[1:10,], caption = "Table 4 - Main species caught in the Parana state, Brazil, 2016 (Source: Fundepag)", booktabs = T) %>%
  kable_styling(latex_options = c("striped", "hold_position"))
  
```

Another important piece of information regarding income generated are the main species exported in Brazil. Data from the Ministry of Development, Industry and Foreign Trade (MDIC) show that the main species exported are lobsters, followed by other fish, swim bladders and pelagic fish (tunas and swordfish) (Table 5). The main species exported, lobsters, are caught exclusively by the artisanal sector, with resources threatened by over exploitation (according to the Ministry of Environment). Because of the high value, the lobster fishery is profitable even when the resource is depleted. Exports of lobsters in Brazil reached staggering US$ 73 million in 2017, with no reported landings of the species in the country. 


```{r}
mdic.data = read.csv("MDIC.csv")

###Rank main species exported - value
exports.value = mdic.data %>% 
  filter(carac == "export" & year == "2017") %>%
  group_by(description) %>%
  summarize(sum(value_uss)) %>%
  rename(Product = description, Value_USS = "sum(value_uss)") %>%
  arrange(desc(Value_USS)) %>%
  transform(Percent = Value_USS/sum(Value_USS)*100) 

  kable(exports.value[1:10,], caption = "Table 5 - Main species exported in 2017, Brazil (Source: MDIC)", booktabs = T) %>%
  kable_styling(latex_options = c("striped", "hold_position"))
```


#4. Domestic Fish Consumption

Using data from FAO and quantity imported and exported, it is possible to estimate the amount of seafood that is consumed in Brazil per year. The estimated consumption of seafood in Brazil is: 760k tonnes from fisheries plus 750k tonnes from aquaculture plus 380k tonnes from imports minus 37k tonnes from exports. This sums to about 1.86 million tonnes of seafood consumed in Brazil. If we divide this number by the number of people (about 200 million) we estimate a consumption of about 9.3 kg/capita/year.

##Top Fisheries by Domestic Consumption

Domestic consumption of fish and seafood in Brazil is estimated to be around 10 kilograms per capita per year (FAO, 2014). Per capita consumption in Brazil is low relive to other countries (average of 18 kg in the world) and the World Health Organization recommendation (12 kg). However, fish and seafood consumption has been steadily increasing from 4.6 kg per capita in 1961 to about 10 kg per capita in 2014 (FAO). Because of Brazil's large population (about 200 million people), this sums up to large amounts of seafood being consumed in the country, although consumption significantly varies across regions of the country. Additionally, estimates from FAO does not account for subsistence consumption of fish, where in some regions of the Amazon can be as high as 169 kg/capita/year (Isaac et al 2015). 

Despite large potential for seafood production from both capture fisheries and aquaculture, Brazil imports large quantities of seafood for domestic consumption. In 2017, Brazil imported more than 380 thousand tonnes of fish and seafood products, while exporting less than 10% of that value (Figure 2). This shows that local production is not meeting seafood demand in the country and increased national production, especially from aquaculture, has great potential to fill this gap.


```{r}
mdic.data = read.csv("MDIC.csv")

mdic.year = mdic.data %>% 
  group_by(year, carac) %>%
  summarize(sum(quantity_kg)) %>%
  rename(prod = "sum(quantity_kg)") 

ggplot(mdic.year, aes(x=year,y=prod,fill=carac)) + 
  geom_bar(stat = "identity", position = position_stack(reverse = TRUE)) + 
  labs(x="Year", y="Quantity (kg)", fill="", caption = "Figure 2 - Seafood and fish imports and exports from Brazil between 2014 to 2017 (Source: MDIC)") +
  ggtheme_plot()

```

Fish and seafood import data shows that the main imported species in 2017 was the sardine followed by salmon, catfish fillet (pagasius) and hake fillet (Table 6). There was a major increase in the import of sardine in the last two years, from 17 thousand tonnes imported in 2014 to 90 thousand tonnes in 2017. This increase is mainly result of the decline in sardine catches from the industrial fleet. According to information from research developed about this fishery, about 100 thousand tonnes of sardine was landed in 2014 followed by a decline to about 20 thousand tonnes in 2017 (Pincinato and Asche 2018). According to the authors, this decline is mainly because of environmental variability in the south of Brazil and "El Niño" events that occurred during that period. Thus, Brazilian sardine and imports form a fully integrated market, which maintains a steady supply of sardines to canary industries year-round. 

```{r}
mdic.data = read.csv("MDIC.csv")

###Rank main species imported - production
imports.quant = mdic.data %>% 
  filter(carac == "import" & year == "2017") %>%
  group_by(description) %>%
  summarize(sum(quantity_kg)) %>%
  rename(Product = description, Quantity = "sum(quantity_kg)") %>%
  arrange(desc(Quantity)) %>%
  transform(Percent = Quantity/sum(Quantity)*100) 

 kable(imports.quant[1:10,], caption = "Table 6 - Main species imported in 2017, Brazil (Source: MDIC)", booktabs = T) %>%
  kable_styling(latex_options = c("striped", "hold_position"))

```

Data from SEAGESP market suggests that the main species consumed in Brazil in 2017 was tilapia, followed by croaker, sardine, salmon, sharks and weakfishes (Table 7). From these species, tilapia and salmon are from aquaculture production and the remaining are from wild fisheries. Only in São Paulo market, 4.2k tonnes of tilapia was commercialized, with a total for all fish and seafood products of 42k tonnes in 2016 and 37k tonnes in 2017.


```{r}
##CEAGESP

ceagesp.data = read.csv("producao_seagesp.csv")

###Rank main species - quantity
ceagesp.prod = ceagesp.data %>% 
  filter(year == "2016") %>%
  group_by(species_english) %>%
  summarize(sum(production_kg)) %>%
  rename(Species = species_english, Quantity = "sum(production_kg)") %>%
  arrange(desc(Quantity)) %>%
  transform(Percent = Quantity/sum(Quantity)*100) 

 kable(ceagesp.prod[1:10,], caption = "Table 7 - Main species commercialized in CEAGESP market, Sao Paulo, Brazil in 2017 (Source: CEAGESP)", booktabs = T) %>%
  kable_styling(latex_options = c("striped", "hold_position"))

```

Data from SIGSIF shows that the main species processed in the country in 2017 was tilapia, followed by tambaqui (Colossoma macropomum), sardine, croaker, tambatinga (hybrid aquaculture species from the Amazon), catfish, salmon and tuna (Table 8). From the top 10 species processed in the country, 5 are from aquaculture production and 5 are from wild caught fisheries. A total of 360k tonnes of fish and seafood products were commercialized by 472 processing and distribution companies. This volume corresponded to about 20% of all estimated seafood consumed in the country (1.8b tonnes).

```{r}
data.sigsif = read.csv("SIGSIF.csv")


###Rank main species comercialized 
domestic.fisheries = data.sigsif %>% 
  filter(origin == "PRODUTOR" & year == "2017") %>%
  group_by(species_english) %>%
  summarize(sum(quantity)) %>%
  rename(Species = species_english, Quantity = "sum(quantity)") %>%
  arrange(desc(Quantity)) %>%
  transform(Percent = Quantity/sum(Quantity)*100)  


  kable(domestic.fisheries[1:10,], caption = "Table 8 - Main species comercialized by seafood industries with SIGSIF certificate in Brazil (Source: Minitry of Agriculture)", booktabs = T) %>%
  kable_styling(latex_options = c("striped", "hold_position"))


```

Consumption of fish and seafood products in Brazil increases significantly during religious dates. Data from SEAGESP suggests that seafood consumption increases during Easter, when other types of meat are not consumed in respect of Christian tradition. For this reason, there is a significant peak in fish and seafood consumption during Easter month (Figure 3).

```{r}

ceagesp.data = read.csv("producao_seagesp.csv")

####graph monthly prodcution
ceagesp.month = ceagesp.data %>% 
  filter(year == "2016") %>%
  group_by(month) %>%
  summarize(sum(production_kg)) %>%
  rename(month = month, prod = "sum(production_kg)") %>% 
  transform(monthabb = month.abb[month])

ggplot(ceagesp.month, aes(x = reorder(monthabb, month), y = prod)) + 
  geom_bar(stat = "identity") + 
  labs(x="Month", y="Quantity (Kg)", caption = "Figure 3 - Quantity of fish and seafood products commercialized per month in 2017 at CEAGESP market (Source: CEAGESP)") + 
  ggtheme_plot()
```

Fish and seafood consumption in Brazil has distinct patterns according to the region of the country. As shown in IBGE data, the top consumers of fish and seafood products in the country are from the north and northeast regions (Figure 4). These regions are the poorest of the country (Figure 5) and have the greatest number of registered fishers (MPA, 2013), especially from the small-scale sector. Since most of the artisanal production is consumed locally, these regions have greater availability of fish and seafood products relative to capitals far from fishing activities. In addition to greater availability of fish and seafood products, the North and Northeast regions of Brazil have, in general, greater tradition of consuming seafood products. Because of extensive areas of coastline, the northeast region has tradition of eating seafood in restaurants close to the ocean, especially in the summer. Similarly, the North region has extensive areas of rivers and coast, being part of the culture to eat fish and seafood products. In other parts of the country, people traditionally eat other types of protein. Brazil has historically been an agriculture and livestock production country, with large tradition of eating barbecue and bovine meat. In many places there are cultural barriers for regularly consuming fish and seafood products. However, with increased aquaculture production, many of these states are now producing more fish and as consequence are starting to eat more fish. In addition, people are increasingly aware of the health benefits derived from eating fish and seafood products. In summary, the north and northeast regions are the greatest consumers of fish and seafood in the country because of the availability and tradition of consuming fish and seafood products. Because these regions are the poorest of the country, fish and seafood products are crucial for food security, especially in coastal and riverine areas. 


```{r}
ibge.consumption = read.csv("IBGE_consumption.csv")
ibge.consumption = ibge.consumption %>%
  transform(kg = consumption*365/1000) 

ggplot(ibge.consumption, aes(x=region,y=kg,fill=protein)) + 
  geom_bar(stat = "identity", position = position_stack(reverse = TRUE)) + 
  labs(x="Region", y="Per capita Consumption (kg/year)", fill="", caption = "Figure 4 - Per capita consumption of protein types per region of Brazil (Source: IBGE)") +
  ggtheme_plot()

```

```{r}
ibge.income = read.csv("IBGE_renda.csv")

income.region = ibge.income %>% 
  filter(type == "region") 

ggplot(income.region, aes(x=name,y=per_capita)) + 
  geom_bar(stat = "identity") + 
  labs(x="Region", y="Monthly Per capita Income (R$)", fill="", caption="Figure 5 - Monthly per capta income by region of Brazil in 2014 (Source: IBGE)") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ggtheme_plot()

```

In the North and Northeast regions, where consumption of seafood is highest, it is important to distinguish resources that are important for food security in coastal and riverine areas. Generally, while the most valuable species are usually sold for profit, smaller and less valuable fish are consumed by fishers and their families. Hundreds of species are consumed in these regions, with very little information on the most important species for food security. 

For those people not involved directly in fishing activity, consumption of fish and seafood products is expensive relative to other protein types (Figure 6). Fish and seafood products are on average three times more expensive than poultry, twice the price of pig meat and 25% more expensive than cow meat (Institute of Agriculture Economics - IEA, 2018). Such difference in price can explain in part the low consumption of fish and seafood products in Brazil, especially during an economic crisis. In many cases, consumption of fish and seafood products are mostly in restaurants as a "fancy" dinner and not a protein for the day-to-day life. 


```{r}
meat.price = data.frame(protein = c("Fish", "Bovine", "Pig", "Poultry"), price = c(31.56, 23.18, 15.24, 10.44)) 

ggplot(meat.price, aes(x = reorder(protein, price), y = price)) + 
  geom_bar(stat = "identity") + 
  labs(x="Protein type", y="Price per Kilogram (R$)", caption = "Figure 6 - Average wholesale price of protein types in the state of Sao Paulo, Brazil (Source: IEA, 2018)") + 
  ggtheme_plot()

```

In addition to elevated prices, frozen seafood products available in supermarkets are many times unreliable and with poor quality. According to the Ministry of Agriculture, seafood fraud was detected in 21% of the fish sold in supermarkets. In addition, a preeminent newspaper article pointed out that 70% of frozen fillets sold in supermarkets had excess water, reaching up to one third of the total weight of the fish. This practice makes the product even more expensive and discourages people from buying seafood products (Folha de SP, 2008). Therefore, for the average consumer, seafood products are expensive and unreliable relative to other protein options. This explains in part low seafood consumption levels in many parts of the country.   

Data on the main species consumed in Brazil suggests that consumers are mostly driven by price, although there are some exceptions. From the top 10 seafood products sold in CEAGESP, 7 where below the 0.5 percentile of prices (Table 9). This indicates that the primary driver of domestic consumption is affordability. However, there are some expeditions, such as salmon, shark and shrimps. Consumption of these species are mainly driven by desirability, supplying high-end markets and restaurants. Species such as salmon and shrimp have high demand because of their distinct flavor and color. Other species, especially white fish can be more easily substituted by other similar fish.  

```{r}
##CEAGESP

ceagesp.data = read.csv("producao_seagesp.csv")

###Rank main species - quantity
ceagesp.prod = ceagesp.data %>% 
  filter(year == "2016") %>%
  group_by(species_english) %>%
  summarize(sum(production_kg), mean(price_rs)) %>%
  rename(Species = species_english, Quantity_kg = "sum(production_kg)", Price_RS = "mean(price_rs)") %>%
  transform(Price_percentile = ecdf(Price_RS)(Price_RS)) %>% 
  arrange(desc(Quantity_kg))

 kable(ceagesp.prod[1:10,], caption = "Table 9 - Price and quantity of main species commercialized in CEAGESP market, Sao Paulo, Brazil in 2017", booktabs = T) %>%
  kable_styling(latex_options = c("striped", "hold_position"))

```

The aquaculture industry in Brazil has been growing rapidly. According the Brazilian fish farming association (PeixeBR), the aquaculture industry is the fastest growing food production sector in the country (PeixeBR, 2018). Brazil has some the largest potential for freshwater aquaculture in the world and will become a significant producer of fish in the next years. Fish farming has been growing around 8% per year in recent years, with a large increase in farming of native fish, especially in the north and middle-west regions. Additionally, Brazil is the fourth largest producer of tilapia in the world, with rapidly growing industries in the south and northeast regions. Another important aquaculture industry in the country is shrimp farming in the northeast region, although it has not been growing the production in recent years. With the expansion of the aquaculture industry in the country, the domestic supply of seafood will increase thus increasing accessibility of fish and seafood product for Brazil's population. In the North region, farmed fish is already affecting the market and reducing prices. Today, a large portion of the fish consumed in the region is produced by aquaculture. Tilapia is already the second most commercialized species in the CEAGESP market in São Paulo and the top species processed by seafood companies in the country. Cachuma (or Tambaqui) is a native species farmed in the north and mid-west regions and is the second most processed fish by seafood industries in Brazil. According to PeixeBR, there was a total production of about 690k tones of fish from aquaculture in 2017 (plus about 80k of shrimp). Thus, aquaculture products are already an important source of seafood for Brazil and has the perspective to grow rapidly in the future.  

#5. Conclusion

Fish and seafood consumption in Brazil is still incipient compared to other parts of the world. However, fisheries resources are important source of food in the poorest regions of the country and therefore are key to food security. The Amazon region, for example, has some of the highest fish consumption in the world, reaching 169kg/capita/year in riverine communities. 

According to the compiled databases, sardines are the most important fish for consumption in Brazil, especially through the canned market. It is an affordable alternative to other protein sources and is consumed all over the country. Other important resources for fish and seafood consumption in the country are: croakers, tilapia, catfish, cachuma, tunas, sharks, salmon, hake, among others. However, these results rely on data mostly from the industrial sector, whcih comprises about 30-50% of catches in the country. The remaining production is from the small-scale sector, where there is basically no information on catches. This sector is characterized by having hundreds of commercially exploited species, where some of  the most important species are crabs, shrimps, lobster, snappers, groupers, mullet, weakfish and snook. 

  
During the study we identified several important data gaps in our analysis. First, there is no national-level fisheries statistics in the country, which hindered our ability to accurately respond some of the questions. Second, there is no data on the fisheries production from the artisanal sector, which accounts for more than half of the seafood production in the coutry. Third, there are very few studies on fish consumption in the country, with the most reliable studies dating back more than eigth years. Fourth, there is very little information on the fisheries supply chain, especially from the artisanal sector. For the industrial sector, many of our databeses did not have information on prices, which made it dificult to estimate the total income generated. 

Regardless of these data gaps, this study was able to gather the best information available. The study highlighted the huge importance of fisheries for income, employment and food security in Brazil. With data from the seafood industry and seafood markets, we now have a better understanding of the main seafood products consumed in the country. With data from IBGE and fisheries licenses, we were able to highlight the importance of fish and seafood products for food security in the poorest parts of the country. Regional landings data helped us better understand the main species caught in some of the most important ports in the country. Import and export data showed the magnitude of the seafood deficit in the country, with a huge potential for aquacuture to fill this gap. 

More importantly, this study highlights the need for better data collection programs and fisheries management policies in Brazil. There is a huge descrepanc between the social importance of this sector and actions of the government to ensure the sustainability of this activity. Urgent action is needed to better understand this important sectorin order to gain the atention of policy makers and achieve fisheries sustainability. 

